---
title: "A guide to the Adventr package"
author: "Andy Field"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A guide to the Adventr packag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

The adventr package contains a series of tutorials that accompany my books [An Adventure in Statistics](https://www.discoveringstatistics.com/books/an-adventure-in-statistics/) and [Discovering statistics using R](https://www.discoveringstatistics.com/books/discovering-statistics-using-r/). These tutorials contain abridged sections from my books so there are some copyright considerations but I offer them under a [Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License](http://creativecommons.org/licenses/by-nc-nd/4.0/), ^[Basically you can use it for teaching and non-profit activities but not meddle with it or claim it as your own work.]

* Who is the package aimed at?
    - Anyone teaching from or reading [An Adventure in Statistics](https://www.discoveringstatistics.com/books/an-adventure-in-statistics/) or [Discovering statistics using R](https://www.discoveringstatistics.com/books/discovering-statistics-using-r/) might find them useful.
* What is covered?
    - The tutorials *do not* teach the background theory (the books do): it is assumed you have either attended my lecture or read the relevant chapter in the aforementioned books (or someone else's)
    - The aim of this tutorial is to augment the theory that you already know by guiding you through fitting linear models using **R** and **RStudio** and asking you questions to test your knowledge along the way.


## Running tutorials

To run a tutorial execute:

```{r, eval = F, include = T}
learnr::run_tutorial("name_of_tutorial", package = "adventr")
```

For example, to run the tutorial on data basics execute:

```{r, eval = F, include = T}

learnr::run_tutorial("r_data_basics", package = "adventr")

```

## A list of tutorials

1. **r_setup**: This tutorial looks at how to download, install, and setup **R** and **RStudio**, although if you're reading this then presumably you've already downloaded, installed and setup **R** amd **RStudio**!
2. **r_data_basics**: This tutorial looks at the basic concepts in **R**, such as installing packages, setting the working directory, getting help, getting data into **R**, creating variables, creating data frames and tibbles, reading in data files, manipulating variables within a tibble/data frame.

